---
title: "S670 Final Project"
author:
- The Fantastic Four
- Erik "Human Torch" Parker
- Emily "Invisible Woman" Rudman
- Vinay "The Thing" Vernekar
- Jervis "Mister Fantastic" Wang
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, warning = FALSE)
```
# Description of Data

__Data Sources__ - Two primary data sources are:   

a) __Catalog.data.gov__ - The data contains details like store/restaurant proximity, food prices, community characteristics etc.  

b) __Factfinder.census.gov__ - Income details (median income) state and county level


__Important Variables__  

__PCT_OBESE_ADULTS10__ - The percentage of obese adults in a county in 2010.

__NATAMEN__ - The Natural Amenities scale, is an index developed by the USDA in 1999 which ranges from 1 to 7 and measures the desirability of a particular location (here a county) based on natural factors.

__FSRPTH07__ - Is the measure for full service restaurants in a county, per 1000 residents, in 2007.

__PCT_DIABETES_ADULTS10__ - The percentage of adults that have diabetes in a county in 2010.

__Above__ - A binary variable that assigns a 1 to a county who is above the national average obesity rate of 26.7% and 0 otherwise.

__Poverty Rate__ - The poverty rate is the ratio of the number of people whose income falls below the poverty line.



```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(GGally)
library(gridExtra)
library(fiftystater)
library(colorplaner)
library(mapproj)
library(viridis)



data <- read.csv("data_Food.csv", header=TRUE)
data <- data[-3144,]
# Read in data and remove last "sanity check" row

data$above <- ifelse(data$PCT_OBESE_ADULTS10<=26.7,0,1)
data$POVRATE<-as.numeric(as.character(data$POVRATE10))
# Make new binomial variable for if a county has an obesity rate above the national average. 26.7 here used as it was the national average for obesity in 2010

model.data <- na.omit(data)
# Remove NAs from the dataset as they screw up the model

model.data$regions <- state.region[match(model.data$State,state.abb)]


# Assign regions to each county based on provided State code

ggpairs(data, columns = c(as.character("NATAMEN"), "POVRATE", "FSRPTH07","PCT_OBESE_ADULTS10", "PCT_DIABETES_ADULTS10", "above"), cardinality_threshold = NULL)
# Plot showing correlations between the final predictor variables chosen.

data$states <- state.name[match(data$State,state.abb)]
# Change state abbreviations to full names for maps

data$states <- tolower(data$states)
# Change state names to all lower case to fit the format expected by mapping package

########## Data for race and poverty ########
health=read.table("Health.csv",header = TRUE,sep=",")

socio=read.table("Socio.csv",header=TRUE,sep=",")

inco=read.table("Income.csv",header=TRUE,sep=",")

# Matching regions with states
inco$regions=state.region[match(inco$State,state.abb)]
inco$regions[which(is.na(inco$regions))]<-"Northeast"

# Merging data sets and removing NA's
Health.race=merge(x=health,y=socio,by=c("FIPS","State","County"),all.x=TRUE)
Health.race.inco=merge(x=Health.race,y=inco,by=c("State"))
maindata=na.omit(Health.race.inco)
maindata$POVRATE10<-as.numeric(as.character(maindata$POVRATE10))

#cor(maindata$PCT_OBESE_ADULTS10,maindata$POVRATE10)
# Selecting required variables
myvar<-c("regions","PCT_NHWHITE10","PCT_NHBLACK10","PCT_HISP10","PCT_NHASIAN10","PCT_NHNA10","PCT_NHPI10")
race_data=maindata[myvar]

### Change data race names to more intutive names 

names(race_data)[2]<-"White"
names(race_data)[3]<-"Black"
names(race_data)[4]<-"Hispanic"
names(race_data)[5]<-"Asian"
names(race_data)[6]<-"Native"
names(race_data)[7]<-"Hawaiian"

#names(race_data)
# Data engineering to get individual counts to support group by
mynew<-gather(race_data,race,value,White:Hawaiian)

pop_contribution= mynew%>%group_by(regions,race)%>%summarise(Pop_contribution=mean(value))


# calculating average obese and diabities region wise
avg_diab=maindata%>%group_by(regions)%>%summarise(Pop_dia=mean(PCT_DIABETES_ADULTS10),Pop_obse=mean(PCT_OBESE_ADULTS10),income=mean(Income))

race_obese_diab=merge(x=pop_contribution,y=avg_diab,by="regions",all.x=TRUE)

myvar1<-c("regions","POVRATE10")
poverty_type=maindata[myvar1]
poverty.wide<-gather(poverty_type,region,value,POVRATE10)
poverty.wide$rate=as.numeric(poverty.wide$value)
#summary(poverty.wide)
poverty_avg<-poverty.wide%>%group_by(regions)%>%summarise(avg_poverty=mean(rate))

national_poverty_avg=15.71 # for 2010
########## Data for race and poverty END  ########

```


```{r, echo= FALSE}
ggplot(race_obese_diab,aes(x=race,y=Pop_contribution,fill=race))+geom_bar(stat="identity")+facet_grid(~regions)+
  geom_hline(aes(yintercept=Pop_dia),colour="#BB0000", linetype="dashed")+
  geom_hline(aes(yintercept=Pop_obse),colour="#000099", linetype="twodash")+
  geom_text(aes( 2, Pop_dia, label = "avg-diabetic",hjust=0,vjust=2), size = 3)+
  geom_text(aes( 2, Pop_obse, label = "avg-obese"), vjust=2,size = 3)+theme(axis.line=element_blank(),axis.text.x=element_blank())+
  ggtitle("Region wise race and health")+
  theme(plot.title = element_text(hjust = 0.5))+theme(legend.position="bottom")
```

__Observations__  
1) Regions have considerable difference in the race diversity  
2) South and West regions have more diversity  
3) South region has more obese rate and diabetes rate   


```{r, echo= FALSE}
avg_inc=inco%>%group_by(regions)%>%summarise(avg_income=mean(Income))
I1=ggplot(avg_inc,aes(x=regions,y=avg_income,fill=regions))+geom_bar(stat="identity")+ggtitle("Region wise avg-median Income")+
  theme(plot.title = element_text(hjust = 0.5))+theme(axis.line=element_blank(),axis.text.x=element_blank())
I2=ggplot(maindata,aes(x=POVRATE10,y=PCT_OBESE_ADULTS10,colour=regions))+geom_point()+
  ggtitle("Poverty and Obesity")+theme(plot.title = element_text(hjust = 0.5))+labs(x="Poverty Rate")

grid.arrange(I1, I2, ncol = 2)

```

__Observations__  
1) South has the lowest avg median income level  
2) Earlier graphs indicate high obesity and diabetes rate in south  
3) The scatter chart clearly indicates some relationship between poverty and obesity  

```{r, echo = FALSE}
ggplot(data, aes(map_id=states)) +  geom_map(aes(fill = NATAMEN), map = fifty_states) + 
  expand_limits(x = fifty_states$long, y = fifty_states$lat) +
  coord_map() +
  scale_x_continuous(breaks = NULL) + 
  scale_y_continuous(breaks = NULL) +
 labs(x = "", y = "", title = "A heat map of Obesity Rate vs Natural Amenities Scale") +
  theme(legend.position = "bottom", 
        panel.background = element_blank()) + aes(fill2 = PCT_OBESE_ADULTS10) + scale_fill_colorplane() +
  theme(legend.position = "right")
# map of NATAMEN and obesity
```

__Observations__  
1) In general there seems to be a regional relationship between both obesity rates and NATAMEN scores.  
2) High NATAMEN is generally matched by low obesity rates and vice versa.  

```{r, echo=FALSE}
ggplot(data, aes(map_id=states)) +  geom_map(aes(fill = NATAMEN), map = fifty_states) + 
  expand_limits(x = fifty_states$long, y = fifty_states$lat) +
  coord_map() +
  scale_x_continuous(breaks = NULL) + 
  scale_y_continuous(breaks = NULL) +
   labs(x = "", y = "", title = "A heat map of Natural Amenities Scale\naveraged by states") +
  theme(legend.position = "bottom", 
        panel.background = element_blank())  +  scale_fill_viridis()
# Map of NATAMEN, averaged by state

ggplot(data, aes(map_id=states)) +  geom_map(aes(fill = PCT_OBESE_ADULTS10), map = fifty_states) + 
  expand_limits(x = fifty_states$long, y = fifty_states$lat) +
  coord_map() +
  scale_x_continuous(breaks = NULL) + 
  scale_y_continuous(breaks = NULL) +
  labs(x = "", y = "", title = "A heat map of Obesity Rate") +
  theme(legend.position = "bottom", 
        panel.background = element_blank()) +  scale_fill_viridis()
# Map of obesity rate

```

__Observations__  
1) NATAMEN scores are generally clusterd by region, with the highest scores seen in Western states and the lowest in midwestern ones.  
2) Likewise, obesity rates are also generally clustered regionally - with lower rates generally in the West and higher ones in the midwest and south.  
3) There are some anomalies, such as the high rate of obesity seen in Washington state despite its high NATAMEN score, and the extremely high rates of obesity seen in the south despite the regions relatively high average NATAMEN scores.  

```{r}
model.ob <- glm(above ~  FSRPTH07 + NATAMEN + POVRATE10 , family = "binomial", data = model.data)

#summary(model.ob)
```

__Observations__  
1) Binomial model predicting if counties are above average obesity from full service resturants/1000 people, NATAMEN, and poverty rate  
2) Residual deviance: when poverty rate is included, reduces by ~200 pts compared to excluding poverty  
3) AIC is also lower when we have poverty rate (better model)  

```{r, echo = FALSE}
model.ob.df <- model.data
model.ob.df$.fitted <- fitted.values(model.ob)
model.ob.df$.resid <- residuals(model.ob, type = "response")
# Pulling out fitted and residual values for residual plot
```


```{r, echo = FALSE}
ggplot(model.ob.df, aes(x = .fitted, y = .resid)) + geom_point() + geom_smooth(method = "loess", method.args = list(degree = 1))+labs(title = "A residual plot of the model")
# Residual plot to check fit of model
```

__Observations__  
1) Interactions were fitted, however none of these other alternative models lead to a significantly better fit or reduction in variance.  
2) Easy to interprete.  
3) Missing values were removed leading to loss of some original explanatory power. 

```{r, echo = FALSE}
ggplot(model.data, aes(x = NATAMEN, y = above)) + geom_point() + geom_jitter(width = 0.2, height = 0.05) + geom_smooth(method = "loess", method.args = list(degree=1))+ labs(title = "Plot of above vs Natural Amenities Scale")
# Plot showing the number of counties above/below average by NATAMEN
```

__Observations__  
1) There is a clear relationship between above and NATAMEN scores.  
2) No counties with a score of 1 are seen to be below average, and no counties with a score of 7 are above average.  
3) After reaching a NATAMEN score of 5, the majority of counties are above average but after this point, there is a quite sudden switch with counties largely being below average for obesity rate.  
4) The majorty of counties included in this analysis are above average for obesity.  

```{r, echo=FALSE}
above.pred <- predict(model.ob, type = "response", newdata = model.data)

above.pred.df <- data.frame(model.ob.df, above.prob = as.vector(above.pred))
# Setting up new dataframe values to get predictions for prediction plots.

ggplot(above.pred.df, aes(x = NATAMEN, y = above.prob, color = regions)) + geom_point() + 
  geom_jitter(width = 0.3, height = 0) + 
  labs(colour = "Region", y = "Probability")+ labs(colour = "Region", y = "Probability", title = "Probability vs NATAMEN colored by region") 
# Probability of being above the national average vs NATAMEN colored by region.  Shows that the west generally has higher NATAMEN and lower prob, NC lowest NATAMEN and high prob, and South high prob moderate NATAMEN.
```

__Observations__  
1) West has highest NATAMEN and lowest probability  
2) North Central lowest NATAMEN, really high probabilities  
3) South moderate NATAMEN, relatively high probability  
4) Northeast very similar to South, with lower NATAMEN overall  


```{r, echo = FALSE}
ggplot(above.pred.df, aes(x = regions, y = above.prob, color = as.factor(NATAMEN))) + geom_point() + 
  geom_jitter(width = 0.3, height = 0) +
  labs(colour = "NATAMEN", y = "Probability", x = "Region", title = "Probability vs region by colored by NATAMEN") 
# Probability vs region, colored by NATAMEN.  Another way to look at previous plot.
```

__Observations__  
1) Another view of previous plot, in general see lower NATAMEN values at high probabilities  
2) West with largest consistent range of probabilities  

```{r, echo = FALSE}
ggplot(above.pred.df, aes(x = NATAMEN, y = above.prob, color = as.numeric(as.character(POVRATE10)))) + geom_point() +
  facet_wrap("regions") + geom_jitter(width = 0.3, height = 0) + 
  scale_color_viridis(option = "inferno") +labs(colour = "Poverty", y = "Probability", x = "NATAMEN", title = "A probability plot of obesity vs NATAMEN colored by poverty rate") 
# Probability vs NATAMEN, colored by poverty rate, and faceted by region.  Shows that the south in general has really high poverty and probability of being above average for obesity.  West seems to be really mixed with poverty rate, but has relatively low probability, maybe because of the really high NATAMEN scores.  Northeast has really high obesity probability as well in general, but generally low poverty.
```

__Observations__   
1) South with generally high poverty and probability, wide range of NATAMEN  
2) West really mixed poverty rate, but generally low probabilities  
3) Northeast low poverty, high probabilities  
4) North Central like South, but lower poverty  
5) In general, seems to be lower probability with lower poverty

```{r, echo = FALSE}
ggplot(above.pred.df, aes(x = as.numeric(as.character(POVRATE10)), y = above.prob, color = as.factor(NATAMEN))) + 
  geom_point() + facet_wrap("regions") + 
  geom_jitter(width = 0.3, height = 0)+ labs(colour = "NATAMEN",x="Poverty Rate", y = "Probability", title = "A probability plot of obesity vs poverty rate colored by NATAMEN")  
# Plot showing probability of above, vs poverty rate, colored by NATAMEN and faceted by region.  Interesting because the West is much different than other regions, and south and north central both show really dramatic relationships of increased probability with increased poverty rate.
```

__Observations__   
1) See clearly increasing probability with increasing poverty in in all regions but west  
2) Also see some horizontal "bands" of probability by NATAMEN  


```{r, echo = FALSE}
ggplot(above.pred.df, aes(x = regions, y = above.prob)) + geom_boxplot() + labs(x="Region", y = "Probability", title = "Summary of probability by region")  

```

__Observations__   
1) Summary of previous figures: see large differences in probabilities by region.  


```{r, echo = FALSE}

#c <- ggplot(above.pred.df, aes(x = CONVSPTH07, y = above.prob, color = regions)) + geom_point() + labs(y = "Probability", x = "Convience Stores / 1000", title = "A probability plot of being obese vs\nConvience Stores coloured by regions") 

ggplot(above.pred.df, aes(x = FSRPTH07, y = above.prob, color = regions)) + geom_point()  + labs(y = "Probability", x = "Full Service Resturants / 1000", title = "A probability plot of obesity full service resturants coloured by region") 

#grid.arrange(c,d,ncol = 2)
```

__Observations__  
1) Possibly some interaction going on? More resturants in West which has lower obesity.  Not resturant number itself driving down rates, just a correlation?  

```{r, echo= FALSE}
ggplot(above.pred.df, aes(x = above, y = NATAMEN, color = regions)) + geom_point() + geom_jitter(width = 0.2, height = .2)+facet_wrap(~State) + labs(title = "A plot of Natural Amenities Scale vs above colored by\nregions and faceted by states")
  
  
```




#Limitations
-We weren't sure which type of diabetes this measures.  
-Didn't report how they got the data.  
-This is an observational study so they had no control over the variables. We just looked for explanatory patterrs.  
-We had to take a a lot of NA and had to take out a bunch of data points ( only 35 states left).  
-The income reported is the median income for each county and has an error margin of +- 5%.  
-We would have liked more categories for the information they had.   


#Conclusion

-We determined FSRPTH07, NATAMEN, and Poverty Rate were the three best variables to predict whether your county would be above and below obesity rate.   
-We made a good predictive model.  
-Don't live in Kentucky.  





